---
title: "QC_male"
format: html
---

# Load dependencies
```{r setup}
#| message: false
#| warning: false

source(here::here("R/library.R"))
load(here::here("data/metadata_new_filtered.rda"))
load(here::here("data/prot_removeoutlier.rda"))
load(here::here("data/df_male.rda"))
```


# Subgroup Male
```{r subgroup sex}
metadata_M <- metadata_new_filtered %>% filter(sex == "M")
df_male <- prot_removeoutlier[, metadata_M$sample_id]
print(metadata_M$sample_id)
print(colnames(df_male))
all(colnames(df_male) == metadata_M$sample_id) # TRUE
dim(df_male) # 8115 proteins, 116 samples


# 1) Add new sample IDs to metadata_M
metadata_M <- metadata_M %>%
  mutate(new_sample_id = paste0("S", seq_len(n())))

# 2) Rename columns in df_male using the new IDs
colnames(df_male) <- metadata_M$new_sample_id
head(colnames(df_male))


metadata_M <- metadata_M %>%
  mutate(Group = paste(diet, strain, sep = "_"))

table(metadata_M$diet) # FFMD-76, LFD-40
table(metadata_M$strain) # J-58, Ntac =58
table(metadata_M$Group) # FFMD_J = 39, LFD_J = 19, FFMD_Ntac = 37, LFD_Ntac = 21

usethis::use_data(df_male, overwrite = T)
usethis::use_data(metadata_M, overwrite = T)
```


# Missing values
```{r missing values}
# Missing values
dim(df_male) # 8115-116
sum(is.na(df_male)) # 68049

# 1. Prepare the binary NA matrix (0 = present, 1 = missing)
missing_matrix <- as.matrix(is.na(df_male)) * 1

annotation_col <- metadata_M %>%
  dplyr::select(new_sample_id, diet, strain, weeks) %>%
  tibble::column_to_rownames(var = "new_sample_id")

# alignment check
stopifnot(all(colnames(missing_matrix) == rownames(annotation_col))) # will error if not aligned

# --- Step 4: Generate the Final Heatmap ---
na_heatmap_male <- pheatmap(
  missing_matrix,                    # The 0/1 data matrix
  annotation_col = annotation_col,   # The annotation data frame with correct rownames
  cluster_rows = FALSE,              # Don't cluster thousands of genes, it's slow and uninformative
  cluster_cols = TRUE,               # Cluster samples to see which are similar in missingness
  show_rownames = FALSE,             # Don't show the gene names, there are too many
  show_colnames = FALSE,             # Hide sample IDs
  color = c("gray95", "black"),      # 0 = Present, 1 = Missing
  legend_breaks = c(0, 1),
  legend_labels = c("Present", "Missing"),
  main = "Pattern of Missing Values Across Samples (Male)"
)

ggsave(file.path("doc", "na_heatmap_male.png"),
       plot = na_heatmap_male, width = 7, height = 6, dpi = 300, bg = "white")

##################################### valid values

valid_value_counts <- colSums(!is.na(df_male))
count_data <- data.frame(column = names(valid_value_counts), count = valid_value_counts)

mean_value <- mean(count_data$count)
n_samples <- length(valid_value_counts)

vv1_male <- ggplot(count_data, aes(x = seq_along(column), y = count)) +
  geom_bar(stat = "identity", fill = "lightblue4") +
  geom_hline(yintercept = mean_value, color = "red", linetype = "dashed", size = 1) +
  geom_text(aes(x = length(column) * 0.8, y = mean_value,
                label = paste0("Mean = ", round(mean_value, 0))),
            vjust = -0.5, color = "red", size = 4) +
  labs(title = paste0("Valid Values per sample, only male (n=", n_samples, ")"),
       x = paste0("Samples (S1–S", n_samples, ")"),
       y = "Valid Values") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank()
  )

ggsave(file.path("doc", "vv1_male.png"),
       plot = vv1_male, width = 7, height = 4, dpi = 300, bg = "white")

```
# Filtering
```{r filtering}
male_log2_filt100 <- PhosR::selectOverallPercent(df_male, 1) # 6119
male_log2_filt70 <- PhosR::selectOverallPercent(df_male, 0.7) # 7358
male_log2_filt <- PhosR::selectOverallPercent(df_male, 0.5) # 7613

usethis::use_data(male_log2_filt, overwrite = T)
```
## Missing values2
```{r}
dim(male_log2_filt)
sum(is.na(male_log2_filt)) # 60704

# 1. Prepare the binary NA matrix (0 = present, 1 = missing)
missing_matrix <- as.matrix(is.na(male_log2_filt)) * 1

annotation_col <- metadata_M %>%
  dplyr::select(new_sample_id, diet, strain, weeks) %>%
  tibble::column_to_rownames(var = "new_sample_id")

# alignment check
stopifnot(all(colnames(missing_matrix) == rownames(annotation_col))) # will error if not aligned

# --- Step 4: Generate the Final Heatmap ---
na_heatmap_male_norm <- pheatmap(
  missing_matrix,                    # The 0/1 data matrix
  annotation_col = annotation_col,   # The annotation data frame with correct rownames
  cluster_rows = FALSE,              # Don't cluster thousands of genes, it's slow and uninformative
  cluster_cols = TRUE,               # Cluster samples to see which are similar in missingness
  show_rownames = FALSE,             # Don't show the gene names, there are too many
  show_colnames = FALSE,              # Hide sample IDs
  color = c("gray95", "black"),       # Use two distinct colors for Present (0) and Missing (1)
  legend_breaks = c(0, 1),            # Specify breaks for the legend
  legend_labels = c("Present", "Missing"), # Label the legend clearly
  main = "Male, filtered_normalized"  # Add a title
)

ggsave(file.path("doc", "na_heatmap_male_norm.png"),
       plot = na_heatmap_male_norm, width = 7, height = 6, dpi = 300, bg = "white")


valid_value_counts <- colSums(!is.na(male_filt50_median))
count_data <- data.frame(column = names(valid_value_counts), count = valid_value_counts)

mean_value <- mean(count_data$count)

vv2_male <- ggplot(count_data, aes(x = seq_along(column), y = count)) +
  geom_bar(stat = "identity", fill = "lightblue4") +
  geom_hline(yintercept = mean_value, color = "red", linetype = "dashed", size = 1) +
  geom_text(aes(x = length(column) * 0.8, y = mean_value,
                label = paste0("Mean = ", round(mean_value, 0))),
            vjust = -0.5, color = "red", size = 4) +
  labs(title = "Valid Values per sample, only male (n=116), filtered",
       x = "Samples (s1-s116)",
       y = "Valid Values") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank()
  )

ggsave(file.path("doc", "vv2_male.png"),
       plot = vv2_male, width = 7, height = 4, dpi = 300, bg = "white")

```

# Distribution
It is roughly symmetric, peaking around log2 intensities of 6-8. there is a slight right tail - totally normal. because of very few abundant proteins dominate. 
Improtantly, it is not super skewed like raw data would be. This confirms data is already lof2 transformed
```{r}
################# density plot #################################################
# Convert data to long format
df_long <- tidyr::pivot_longer(male_log2_filt, 
                               cols = everything(), 
                               names_to = "new_sample_id", 
                               values_to = "intensity")

density_plot_male <- ggplot(df_long, aes(x = intensity, color = new_sample_id)) +
  geom_density() +
  labs(title = "Density Distribution, male", x = "Intensity", y = "Density") +
  theme_minimal(base_size = 14) +
  theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.position = "none"
  )

density_plot_male
plotly::ggplotly(density_plot_male)

ggsave(file.path("doc", "density_plot_male.png"), plot = density_plot_male, width = 5, height = 3, dpi = 300, bg = "white")


################# boxplot ######################################################
# base R boxplot
png(file.path("doc", "Boxplot_male.png"), width = 13, height = 6, units = "in", res = 300, bg = "white")
boxplot(male_log2_filt,
        las = 2,                # rotate x-axis labels
        outline = FALSE,        # hide extreme outliers for clarity
        col = "lightblue",
        main = "Male- before normalization",
        ylab = "Log2 intensity")
dev.off()

```


# Normalization median-centering, PhosR
```{r}
male_log2_filt_median <- medianScaling(male_log2_filt[1:ncol(male_log2_filt)], scale = TRUE)
male_log2_filt_median <- as.data.frame(male_log2_filt_median)
usethis::use_data(male_log2_filt_median, overwrite = T)


############################### boxplot ########################################
png(file.path("doc", "Boxplot_male_norm.png"), width = 13, height = 6, units = "in", res = 300, bg = "white")
boxplot(male_log2_filt_median,
        las = 2,                # rotate x-axis labels
        outline = FALSE,        # hide extreme outliers for clarity
        col = "lightgreen",
        main = "After median centering normalization",
        ylab = "Log2 intensity")
dev.off()


############## density plot
# Convert data to long format
df_long <- tidyr::pivot_longer(male_log2_filt_median, 
                               cols = everything(), 
                               names_to = "new_sample_id", 
                               values_to = "intensity")

density_plot_male_norm <- ggplot(df_long, aes(x = intensity, color = new_sample_id)) +
  geom_density() +
  labs(title = "After normalization, male", x = "Intensity", y = "Density") +
  theme_minimal(base_size = 14) +
  theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.position = "none"
  )

density_plot_male_norm
plotly::ggplotly(density_plot_male_norm)

ggsave(file.path("doc", "density_plot_male_norm.png"), plot = density_plot_male_norm, width = 5, height = 3, dpi = 300, bg = "white")

```

## Median absolute deviation
```{r}
# Calculate per-sample MAD before normalization
mad_df_before <- male_log2_filt %>%
  summarise(across(everything(), ~ mad(., na.rm = TRUE))) %>%
  pivot_longer(cols = everything(), names_to = "Sample", values_to = "MAD") %>%
  mutate(Type = "Before Median Centering")

# Calculate per-sample MAD after median centering
mad_df_after <- male_log2_filt_median %>%
  summarise(across(everything(), ~ mad(., na.rm = TRUE))) %>%
  pivot_longer(cols = everything(), names_to = "Sample", values_to = "MAD") %>%
  mutate(Type = "After Median Centering")

# Combine results
mad_compare <- bind_rows(mad_df_before, mad_df_after)

# Plot comparison
MAD_male <- ggplot(mad_compare, aes(x = Sample, y = MAD, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),   # Hide sample labels (too many to display)
    axis.ticks.x = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  labs(
    title = "MAD per Sample: Before vs After Median Centering",
    y = "Median Absolute Deviation (MAD)",
    x = "Sample"
  )

# Save figure
ggsave(
  filename = file.path("doc", "MAD_male.png"),
  plot = MAD_male,
  width = 6, height = 4, dpi = 300, bg = "white"
)

```


## MAD by groups
```{r}
mad_group_stats <- mad_compare %>%
  left_join(
    metadata_M %>% dplyr::select(new_sample_id, diet, strain),
    by = c("Sample" = "new_sample_id")
  ) %>%
  group_by(diet, strain, Type) %>%
  summarise(
    MAD_mean = mean(MAD, na.rm = TRUE),
    MAD_sd   = sd(MAD, na.rm = TRUE),
    n        = sum(!is.na(MAD)),
    MAD_se   = MAD_sd / sqrt(n),
    .groups  = "drop"
  ) %>%
  mutate(Group = interaction(diet, strain, sep = " · ", drop = TRUE))

p_mad_group_se <- ggplot(mad_group_stats,
                         aes(x = Group, y = MAD_mean, fill = Type)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.7) +
  geom_errorbar(
    aes(ymin = MAD_mean - MAD_se, ymax = MAD_mean + MAD_se),
    position = position_dodge(width = 0.85), width = 0.2
  ) +
  theme_bw() +
  labs(
    title = "Mean MAD by Group (±SE): Before vs After Median Centering",
    x = "Diet · Strain",
    y = "Mean MAD"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    legend.title = element_blank()
  )

ggsave(
  filename = file.path("doc", "MAD_group_means_se.png"),
  plot = p_mad_group_se, width = 7, height = 4.5, dpi = 300, bg = "white"
)

```


## All 3 boxplots
```{r}
# -------------------------------
# 1) Extract numeric intensity data only
expr_mat <- as.matrix(male_log2_filt)

# -------------------------------
# 2) Raw (no normalization)
raw_long <- as.data.frame(expr_mat) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Sample",
    values_to = "Intensity"
  ) %>%
  mutate(Normalization = "Raw")

# -------------------------------
# 3) Median-centered (scale = FALSE)
med_centered <- medianScaling(expr_mat, scale = FALSE) %>%
  as.data.frame() %>%
  pivot_longer(
    cols = everything(),
    names_to = "Sample",
    values_to = "Intensity"
  ) %>%
  mutate(Normalization = "Median-centered (scale=FALSE)")

# -------------------------------
# 4) Median-centered + scaled (scale = TRUE)
med_scaled <- medianScaling(expr_mat, scale = TRUE) %>%
  as.data.frame() %>%
  pivot_longer(
    cols = everything(),
    names_to = "Sample",
    values_to = "Intensity"
  ) %>%
  mutate(Normalization = "Median-centered + scaled (scale=TRUE)")

# -------------------------------
# 5) Combine all three
df_long <- bind_rows(raw_long, med_centered, med_scaled) %>%
  mutate(
    Normalization = factor(
      Normalization,
      levels = c(
        "Raw",
        "Median-centered (scale=FALSE)",
        "Median-centered + scaled (scale=TRUE)"
      )
    )
  )

# -------------------------------
# 6) Plot
male_boxplots <- ggplot(df_long, aes(x = Sample, y = Intensity, fill = Normalization)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~Normalization, ncol = 1, scales = "free_y") +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  ) +
  labs(
    title = "Raw vs Median-centered vs Median-centered + Scaled",
    x = "Samples",
    y = "Log2 intensity"
  )


ggsave(file.path("doc", "male_boxplots.png"), plot = male_boxplots, width = 13, height = 12, dpi = 300, bg = "white")
```


# PCA function
```{r}
run_pca_plots <- function(data_matrix, metadata, title_prefix, save_prefix,
                          color_var = "diet", shape_var = "weeks", palette = "Set1") {

  # ---- Step 1: Handle missing values ---------------------------------------
  n_original <- nrow(data_matrix)
  df_nona    <- na.omit(data_matrix)
  n_nona     <- nrow(df_nona)

  # ---- Step 2: PCA ----------------------------------------------------------
  pca_nona      <- prcomp(t(df_nona), scale. = TRUE)
  pca_var       <- pca_nona$sdev^2
  pca_var_perc  <- round(pca_var / sum(pca_var) * 100, 1)

  # ---- Step 3: Merge PCA results with metadata ------------------------------
  pca_results <- as.data.frame(pca_nona$x) %>%
    tibble::rownames_to_column("new_sample_id")

  # Keep the intent of a sanity check without changing behavior
  stopifnot("new_sample_id" %in% names(pca_results))

  pca_data <- pca_results %>%
    dplyr::select(new_sample_id, PC1, PC2, PC3, PC4) %>%
    dplyr::inner_join(metadata, by = "new_sample_id") %>%
    dplyr::mutate(
      weeks = as.factor(weeks),
      hover_text = paste0(
        "Sample: ", sample_id,
        "<br>Diet: ", diet,
        "<br>Strain: ", strain,
        "<br>Weeks: ", weeks
      )
    )

  # ---- Step 4: Base plot ----------------------------------------------------
  p_base <- ggplot(
    pca_data,
    aes_string(x = "PC1", y = "PC2", color = color_var, shape = shape_var, text = "hover_text")
  ) +
    geom_point(size = 3.5, alpha = 0.7) +
    scale_color_brewer(palette = palette) +
    labs(
      title    = paste0(title_prefix, " (n=", ncol(data_matrix), ")"),
      subtitle = paste("based on", n_nona, "proteins out of", n_original),
      x = paste0("PC1 (", pca_var_perc[1], "%)"),
      y = paste0("PC2 (", pca_var_perc[2], "%)")
    ) +
    theme_bw(base_size = 14) +
    theme(
      plot.title    = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
    )

  # ---- Save the plot --------------------------------------------------------
  ggsave(
    file.path("doc", paste0(save_prefix, "_PC1_PC2.png")),
    p_base, width = 8, height = 6, dpi = 300, bg = "white"
  )

  # ---- Return everything needed --------------------------------------------
  return(list(
    pca_data       = pca_data,
    pca_var_perc   = pca_var_perc,
    plot_PC1_PC2   = p_base
  ))
}


before <- run_pca_plots(
  male_log2_filt, metadata_M,
  title_prefix = "PCA-Male (Before Norm)",
  save_prefix = "pca_male_before",
  color_var = "prep_day_batch", 
  shape_var = "diet"
)

after <- run_pca_plots(
  male_log2_filt_median, metadata_M,
  title_prefix = "PCA-Male (After Norm)",
  save_prefix  = "pca_male_after",
  color_var    = "prep_day_batch",
  shape_var    = "diet"
)


combined_pc <- bind_rows(
  before$pca_data %>%
    mutate(stage = paste0(
      "Before (PC1=", before$pca_var_perc[1],
      "%, PC2=", before$pca_var_perc[2], "%)"
    )),
  after$pca_data %>%
    mutate(stage = paste0(
      "After (PC1=", after$pca_var_perc[1],
      "%, PC2=", after$pca_var_perc[2], "%)"
    ))
)


ggplot(combined_pc, aes(PC1, PC2, color = prep_day_batch, shape = diet)) +
  geom_point(size = 3) +
  facet_wrap(~stage) +
  theme_bw() +
  labs(
    title = "PCA Before vs After Normalization",
    x = "PC1",
    y = "PC2"
  )

```


# PCA1 before normalization
```{r}
dim(male_log2_filt) # 8115-116
df_nona <- na.omit(male_log2_filt) # 6119

n_original <- nrow(male_log2_filt)
n_nona <- nrow(df_nona)

# transpose data 
pca_nona <- prcomp(t(df_nona), scale = T)

# quick scatter plot
factoextra::fviz_pca_ind(pca_nona) 
plot(pca_nona$x[,1], pca_nona$x[,2])

# Eigenvalues (variance explained by each PC): standard deviations squared.
pca_var <- pca_nona$sdev^2 
pca_var_perc <- round(pca_var/sum(pca_var)*100, digits = 1) #percentage of each component
fviz_eig(pca_nona, addlabels = T)  #scree plot 

pca_results <- as.data.frame(pca_nona$x) %>%
  tibble::rownames_to_column("new_sample_id")  # keep sample IDs

# double check the alignment
all(pca_results$new_sample_id %in% metadata_M$new_sample_id)
setdiff(pca_results$new_sample_id, metadata_M$new_sample_id)

pca_data_M <- pca_results %>%
  dplyr::select(new_sample_id, PC1, PC2, PC3, PC4) %>% 
  inner_join(metadata_M, by = "new_sample_id") %>%
  mutate(
    weeks = as.factor(weeks),
    hover_text = paste0(
      "Sample: ", sample_id,
      "<br>Diet: ", diet,
      "<br>Strain: ", strain,
      "<br>Weeks: ", weeks
    )
  )


# day batch
pca_M_day <- ggplot(pca_data_M, aes(x = PC1, y = PC2, color = prep_day_batch, shape = diet)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  #stat_ellipse(aes(group = as.factor(cage_id)), type = "t", level = 0.95, geom = "path", show.legend = FALSE) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCA-Male: prep day batch (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_day)
plotly::ggplotly(pca_M_day)

ggsave(file.path("doc", "pca_M_day.png"), plot = pca_M_day, width = 8, height = 6, dpi = 300, bg = "white")


#beatbox batch
pca_M_beatbox <- ggplot(pca_data_M, aes(x = PC1, y = PC2, color = beatbox_batch, shape = prep_day_batch)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  #stat_ellipse(aes(group = as.factor(cage_id)), type = "t", level = 0.95, geom = "path", show.legend = FALSE) +
  #scale_color_brewer(palette = "Set3") +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "PCA-Male: beatbox batch (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_beatbox)
plotly::ggplotly(pca_M_beatbox)

ggsave(file.path("doc", "pca_M_beatbox.png"), plot = pca_M_beatbox, width = 8, height = 6, dpi = 300, bg = "white")





##################### Diet, strain ##################
pca_M_diet <- ggplot(pca_data_M, aes(x = PC1, y = PC2, color = diet, shape = weeks, text = hover_text)) + 
  geom_point(size = 3.5, alpha = 0.7) + 
  stat_ellipse(aes(group = diet, fill = diet), geom = "path", show.legend = FALSE) + 
  scale_color_manual(values = c("LFD" = "#21908CFF", "FFMD" = "#D55E00" )) +
  labs(
    title = "PCA-Male: diet & weeks (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_diet)
plotly::ggplotly(pca_M_diet)

ggsave(file.path("doc", "pca_M_diet.png"), plot = pca_M_diet, width = 7, height = 6, dpi = 300, bg = "white")


#####################
pca_M_strain <- ggplot(pca_data_M, aes(x = PC1, y = PC2, color = strain, shape = weeks, text = hover_text)) + 
  geom_point(size = 3.5, alpha = 0.7) + 
  stat_ellipse(aes(group = strain, fill = strain), geom = "path", show.legend = FALSE) + 
  scale_color_manual(values = c("Ntac" = "#AA3377", "J" = "#CCBB44" )) +
  labs(
    title = "PCA-Male: strain & weeks (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_strain)
plotly::ggplotly(pca_M_strain)

ggsave(file.path("doc", "pca_M_strain.png"), plot = pca_M_strain, width = 7, height = 6, dpi = 300, bg = "white")


#####################
pca_M_weeks <- ggplot(pca_data_M, aes(x = PC1, y = PC2, color = weeks, shape = strain)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "PCA-Male: weeks & strain (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_weeks)
plotly::ggplotly(pca_M_weeks)

ggsave(file.path("doc", "pca_M_weeks.png"), plot = pca_M_weeks, width = 7, height = 6, dpi = 300, bg = "white")

```






# PCA2 after normalization
```{r}
dim(male_log2_filt_median) # 7612-116
df_nona <- na.omit(male_log2_filt_median) # 6168

n_original <- nrow(male_log2_filt_median)
n_nona <- nrow(df_nona)

# transpose data 
pca_nona <- prcomp(t(df_nona), scale = T)

# quick scatter plot
factoextra::fviz_pca_ind(pca_nona) 
plot(pca_nona$x[,1], pca_nona$x[,2])

# Eigenvalues (variance explained by each PC): standard deviations squared.
pca_var <- pca_nona$sdev^2 
pca_var_perc <- round(pca_var/sum(pca_var)*100, digits = 1) #percentage of each component
fviz_eig(pca_nona, addlabels = T)  #scree plot 

pca_results <- as.data.frame(pca_nona$x) %>%
  tibble::rownames_to_column("new_sample_id")  # keep sample IDs

# double check the alignment
all(pca_results$new_sample_id %in% metadata_M$new_sample_id)


pca_data_M_norm <- pca_results %>%
  dplyr::select(new_sample_id, PC1, PC2, PC3, PC4) %>% 
  inner_join(metadata_M, by = "new_sample_id") %>%
  mutate(
    weeks = as.factor(weeks),
    hover_text = paste0(
      "Sample: ", sample_id,
      "<br>Diet: ", diet,
      "<br>Strain: ", strain,
      "<br>Weeks: ", weeks
    )
  )




# day batch
pca_M_day_norm <- ggplot(pca_data_M_norm, aes(x = PC1, y = PC2, color = prep_day_batch, shape = diet)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  #stat_ellipse(aes(group = as.factor(cage_id)), type = "t", level = 0.95, geom = "path", show.legend = FALSE) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCA-Male-norm: prep day batch (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_day_norm )
plotly::ggplotly(pca_M_day_norm )

ggsave(file.path("doc", "pca_M_day_norm .png"), plot = pca_M_day_norm , width = 8, height = 6, dpi = 300, bg = "white")


#beatbox batch
pca_M_beatbox_norm  <- ggplot(pca_data_M_norm, aes(x = PC1, y = PC2, color = beatbox_batch, shape = prep_day_batch)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  #stat_ellipse(aes(group = as.factor(cage_id)), type = "t", level = 0.95, geom = "path", show.legend = FALSE) +
  #scale_color_brewer(palette = "Set3") +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "PCA-Male-norm: beatbox batch (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_beatbox_norm )
plotly::ggplotly(pca_M_beatbox_norm )

ggsave(file.path("doc", "pca_M_beatbox_norm .png"), plot = pca_M_beatbox_norm , width = 8, height = 6, dpi = 300, bg = "white")





##################### Diet, strain ##################
pca_M_diet_norm  <- ggplot(pca_data_M_norm, aes(x = PC1, y = PC2, color = diet, shape = weeks, text = hover_text)) + 
  geom_point(size = 3.5, alpha = 0.7) + 
  stat_ellipse(aes(group = diet, fill = diet), geom = "path", show.legend = FALSE) + 
  scale_color_manual(values = c("LFD" = "#21908CFF", "FFMD" = "#D55E00" )) +
  labs(
    title = "PCA-Male-norm: diet & weeks (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_diet_norm )
plotly::ggplotly(pca_M_diet_norm )

ggsave(file.path("doc", "pca_M_diet_norm .png"), plot = pca_M_diet_norm , width = 7, height = 6, dpi = 300, bg = "white")


#####################
pca_M_strain_norm  <- ggplot(pca_data_M_norm, aes(x = PC1, y = PC2, color = strain, shape = weeks, text = hover_text)) + 
  geom_point(size = 3.5, alpha = 0.7) + 
  stat_ellipse(aes(group = strain, fill = strain), geom = "path", show.legend = FALSE) + 
  scale_color_manual(values = c("Ntac" = "#AA3377", "J" = "#CCBB44" )) +
  labs(
    title = "PCA-Male-norm: strain & weeks (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_strain_norm )
plotly::ggplotly(pca_M_strain_norm )

ggsave(file.path("doc", "pca_M_strain_norm .png"), plot = pca_M_strain_norm , width = 7, height = 6, dpi = 300, bg = "white")


#####################
pca_M_weeks_norm  <- ggplot(pca_data_M_norm , aes(x = PC1, y = PC2, color = weeks, shape = strain)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "PCA-Male-norm: weeks & strain (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_weeks_norm )
plotly::ggplotly(pca_M_weeks_norm )

ggsave(file.path("doc", "pca_M_weeks_norm .png"), plot = pca_M_weeks_norm , width = 7, height = 6, dpi = 300, bg = "white")

```


# Missing proteins
```{r missing proteins}
data_long <- male_filt50_median %>%
  rownames_to_column(var = "PG.Genes") %>%
  pivot_longer(
    cols = -PG.Genes,              # Pivot all columns except gene names
    names_to = "new_sample_id",    # New column for sample names
    values_to = "log2_abundance"   # New column for abundance values
  )

protein_missing_summary <- data_long %>%
  group_by(PG.Genes) %>%
  summarize(
    n_total_samples = n(),                     # Total number of samples
    n_missing = sum(is.na(log2_abundance)),    # Count NAs per protein
    n_present = sum(!is.na(log2_abundance))    # Count present values
  ) %>%
  mutate(
    percent_missing = (n_missing / n_total_samples) * 100,
    percent_present = (n_present / n_total_samples) * 100
  ) %>%
  arrange(desc(percent_missing))

################################################################################
status_matrix <- male_filt50_median %>%
  rownames_to_column(var = "PG.Genes") %>%
  mutate(across(
    .cols = -PG.Genes,
    .fns = ~ if_else(is.na(.), "Missing", "Present")
  ))

# Example: check a specific protein
status_matrix %>%
  filter(PG.Genes == "Lime1")

# Example: see all proteins for a specific sample, "S5"
status_matrix %>%
  dplyr::select(PG.Genes, S5)

######################################
ggplot(protein_missing_summary, aes(x = percent_missing)) +
  geom_histogram(binwidth = 2, fill = "#56B4E9", color = "white", alpha = 0.85) +
  labs(title = "Distribution of Missingness Across Proteins (Male)",
       x = "Percent Missing (per protein)",
       y = "Protein Count") +
  theme_minimal(base_size = 14)

top_missing <- protein_missing_summary %>% 
  slice_max(order_by = percent_missing, n = 10)

sample_missing_summary <- data_long %>%
  group_by(new_sample_id) %>%
  summarize(n_total_proteins = n(),
            n_missing = sum(is.na(log2_abundance)),
            percent_missing = 100 * n_missing / n_total_proteins) %>%
  arrange(desc(percent_missing))

###################
# Count proteins with 0% missingness
n_zero_missing <- sum(protein_missing_summary$percent_missing == 0, na.rm = TRUE)

missigness_plot_male <- ggplot(protein_missing_summary, aes(x = percent_missing)) +
  geom_histogram(
    binwidth = 2,
    fill = viridis(1, option = "D"),
    color = "grey90",
    alpha = 0.95,
    boundary = 0
  ) +
  labs(
    title = "Distribution of Missingness, Male (n=116)",
    x = "Percentage of Missing Values per Protein",
    y = "Number of Proteins"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5, margin = margin(b = 6)),
    plot.subtitle = element_text(hjust = 0.5),
    axis.title.x = element_text(face = "bold", size = 13),
    axis.title.y = element_text(face = "bold", size = 13),
    legend.position = "none"
  ) +
  annotate(
    "text",
    x = 10,  # near the first few bins
    y = max(table(cut(protein_missing_summary$percent_missing, breaks = seq(0, 100, by = 2)))) * 0.9,
    label = paste("Proteins with 0% missingness:", n_zero_missing),
    hjust = 0,
    size = 5,
    fontface = "bold"
  )

print(missigness_plot_male)

ggsave(file.path("doc", "missigness_plot_male.png"),
       plot = missigness_plot_male, width = 6, height = 5, dpi = 300, bg = "white")

```


# Batchcorr beatbox
```{r}
setequal(colnames(male_log2_filt_median), metadata_M$new_sample_id) # TRUE
all(colnames(male_log2_filt_median) == metadata_M$new_sample_id)    # TRUE

male_filt_median_batchcorr <- limma::removeBatchEffect(male_log2_filt_median, batch = metadata_M$beatbox_batch)
male_filt_median_batchcorr <- as.data.frame(male_filt_median_batchcorr)
usethis::use_data(male_filt_median_batchcorr, overwrite = TRUE)


male_filt_median_batchcorr_day <- limma::removeBatchEffect( male_log2_filt_median, batch = metadata_M$prep_day_batch)
male_filt_median_batchcorr_day <- as.data.frame(male_filt_median_batchcorr_day)
usethis::use_data(male_filt_median_batchcorr_day, overwrite = TRUE)
```


# PCA3 beatbox_batchcorr
```{r}
dim(male_filt_median_batchcorr) # 7642
df_nona <- na.omit(male_filt_median_batchcorr) # 6354

n_original <- nrow(male_filt_median_batchcorr)
n_nona <- nrow(df_nona)

# transpose data 
pca_nona <- prcomp(t(df_nona), scale = TRUE)
# quick scatter plot
factoextra::fviz_pca_ind(pca_nona) 
plot(pca_nona$x[,1], pca_nona$x[,2])

# Eigenvalues (variance explained by each PC): standard deviations squared.
pca_var <- pca_nona$sdev^2 
pca_var_perc <- round(pca_var / sum(pca_var) * 100, digits = 1) # percentage of each component
fviz_eig(pca_nona, addlabels = TRUE)  # scree plot 

pca_results <- as.data.frame(pca_nona$x) %>%
  tibble::rownames_to_column("new_sample_id")  # keep sample IDs

pca_data <- pca_results %>%
  dplyr::select(new_sample_id, PC1, PC2, PC3, PC4) %>% 
  inner_join(metadata_M, by = "new_sample_id") %>%
  mutate(
    weeks = as.factor(weeks),
    hover_text = paste0(
      "Sample: ", sample_id,
      "<br>Diet: ", diet,
      "<br>Strain: ", strain,
      "<br>Weeks: ", weeks
    )
  )

# Inner join keeps only samples present in both pca_results and metadata.
# Should be all TRUE / empty sets
all(pca_results$new_sample_id %in% metadata_M$new_sample_id)





pca_M_day_batchcorr <- ggplot(pca_data, aes(x = PC1, y = PC2, color = prep_day_batch, shape = diet)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  # stat_ellipse(aes(group = as.factor(cage_id)), type = "t", level = 0.95, geom = "path", show.legend = FALSE) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCA-Male batchcorr: prep day batch (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_day_batchcorr)
plotly::ggplotly(pca_M_day_batchcorr)

ggsave(file.path("doc", "pca_M_day_batchcorr.png"),
       plot = pca_M_day_batchcorr, width = 8, height = 6, dpi = 300, bg = "white")


#########################
pca_M_beatbox_batchcorr <- ggplot(pca_data, aes(x = PC1, y = PC2, color = beatbox_batch, shape = prep_day_batch)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  # stat_ellipse(aes(group = as.factor(cage_id)), type = "t", level = 0.95, geom = "path", show.legend = FALSE) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "PCA-Male batchcorr: beatbox batch (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_beatbox_batchcorr)
plotly::ggplotly(pca_M_beatbox_batchcorr)

ggsave(file.path("doc", "pca_M_beatbox_batchcorr.png"),
       plot = pca_M_beatbox_batchcorr, width = 8, height = 6, dpi = 300, bg = "white")

pca_M_diet_week_batchcorr <- ggplot(pca_data, aes(x = PC1, y = PC2, color = diet, shape = weeks, text = hover_text)) +
  geom_point(size = 3.5, alpha = 0.7) +
  stat_ellipse(aes(group = diet, fill = diet), geom = "path", show.legend = FALSE) +
  scale_color_manual(values = c("LFD" = "#21908CFF", "FFMD" = "#D55E00" )) +
  labs(
    title = "PCA-Male batchcorr: diet & weeks (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_diet_week_batchcorr)
plotly::ggplotly(pca_M_diet_week_batchcorr)

ggsave(file.path("doc", "pca_M_diet_week_batchcorr.png"), 
       plot = pca_M_diet_week_batchcorr, width = 7, height = 6, dpi = 300, bg = "white")


#####################
pca_M_strain_weeks_batchcorr <- ggplot(pca_data, aes(x = PC1, y = PC2, color = strain, shape = weeks, text = hover_text)) +
  geom_point(size = 3.5, alpha = 0.7) +
  stat_ellipse(aes(group = strain, fill = strain), geom = "path", show.legend = FALSE) +
  scale_color_manual(values = c("Ntac" = "#AA3377", "J" = "#CCBB44" )) +
  labs(
    title = "PCA-Male batchcorr: strain & weeks (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_strain_weeks_batchcorr)
plotly::ggplotly(pca_M_strain_weeks_batchcorr)

ggsave(file.path("doc", "pca_M_strain_weeks_batchcorr.png"), 
       plot = pca_M_strain_weeks_batchcorr, width = 7, height = 6, dpi = 300, bg = "white")


#####################
pca_M_weeks_batchcorr <- ggplot(pca_data, aes(x = PC1, y = PC2, color = weeks, shape = strain)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  #stat_ellipse(aes(group = as.factor(cage_id)), type = "t", level = 0.95, geom = "path", show.legend = FALSE) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "PCA-Male batchcorr: weeks & strain (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_weeks_batchcorr)
plotly::ggplotly(pca_M_weeks_batchcorr)

ggsave(file.path("doc", "pca_M_weeks_batchcorr.png"), 
       plot = pca_M_weeks_batchcorr, width = 7, height = 6, dpi = 300, bg = "white")
```



# PCA4 day_batchcorr
```{r}
dim(male_filt_median_batchcorr_day) # 7642
df_nona <- na.omit(male_filt_median_batchcorr_day) # 6354

n_original <- nrow(male_filt_median_batchcorr_day)
n_nona <- nrow(df_nona)

# transpose data 
pca_nona <- prcomp(t(df_nona), scale = TRUE)
# quick scatter plot
factoextra::fviz_pca_ind(pca_nona) 
plot(pca_nona$x[,1], pca_nona$x[,2])

# Eigenvalues (variance explained by each PC): standard deviations squared.
pca_var <- pca_nona$sdev^2 
pca_var_perc <- round(pca_var / sum(pca_var) * 100, digits = 1) # percentage of each component
fviz_eig(pca_nona, addlabels = TRUE)  # scree plot 

pca_results <- as.data.frame(pca_nona$x) %>%
  tibble::rownames_to_column("new_sample_id")  # keep sample IDs

pca_data <- pca_results %>%
  dplyr::select(new_sample_id, PC1, PC2, PC3, PC4) %>% 
  inner_join(metadata_M, by = "new_sample_id") %>%
  mutate(
    weeks = as.factor(weeks),
    hover_text = paste0(
      "Sample: ", sample_id,
      "<br>Diet: ", diet,
      "<br>Strain: ", strain,
      "<br>Weeks: ", weeks
    )
  )

# Inner join keeps only samples present in both pca_results and metadata.
# Should be all TRUE / empty sets
all(pca_results$new_sample_id %in% metadata_M$new_sample_id)





pca_M_day_batchcorrday <- ggplot(pca_data, aes(x = PC1, y = PC2, color = prep_day_batch, shape = diet)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  # stat_ellipse(aes(group = as.factor(cage_id)), type = "t", level = 0.95, geom = "path", show.legend = FALSE) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCA-Male batchcorrday: prep day batch (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_day_batchcorrday)
#plotly::ggplotly(pca_M_day_batchcorrday)

ggsave(file.path("doc", "pca_M_day_batchcorrday.png"),
       plot = pca_M_day_batchcorrday, width = 8, height = 6, dpi = 300, bg = "white")


#########################
pca_M_beatbox_batchcorrday <- ggplot(pca_data, aes(x = PC1, y = PC2, color = beatbox_batch, shape = prep_day_batch)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  # stat_ellipse(aes(group = as.factor(cage_id)), type = "t", level = 0.95, geom = "path", show.legend = FALSE) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "PCA-Male batchcorrday: beatbox batch (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_beatbox_batchcorrday)
#plotly::ggplotly(pca_M_beatbox_batchcorrday)

ggsave(file.path("doc", "pca_M_beatbox_batchcorrday.png"),
       plot = pca_M_beatbox_batchcorrday, width = 8, height = 6, dpi = 300, bg = "white")




pca_M_diet_week_batchcorrday <- ggplot(pca_data, aes(x = PC1, y = PC2, color = diet, shape = weeks, text = hover_text)) +
  geom_point(size = 3.5, alpha = 0.7) +
  stat_ellipse(aes(group = diet, fill = diet), geom = "path", show.legend = FALSE) +
  scale_color_manual(values = c("LFD" = "#21908CFF", "FFMD" = "#D55E00" )) +
  labs(
    title = "PCA-Male batchcorrday: diet & weeks (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_diet_week_batchcorrday)
plotly::ggplotly(pca_M_diet_week_batchcorrday)

ggsave(file.path("doc", "pca_M_diet_week_batchcorrday.png"), 
       plot = pca_M_diet_week_batchcorrday, width = 7, height = 6, dpi = 300, bg = "white")


#####################
pca_M_strain_weeks_batchcorrday <- ggplot(pca_data, aes(x = PC1, y = PC2, color = strain, shape = weeks, text = hover_text)) +
  geom_point(size = 3.5, alpha = 0.7) +
  stat_ellipse(aes(group = strain, fill = strain), geom = "path", show.legend = FALSE) +
  scale_color_manual(values = c("Ntac" = "#AA3377", "J" = "#CCBB44" )) +
  labs(
    title = "PCA-Male batchcorrday: strain & weeks (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_strain_weeks_batchcorrday)
plotly::ggplotly(pca_M_strain_weeks_batchcorrday)

ggsave(file.path("doc", "pca_M_strain_weeks_batchcorrday.png"), 
       plot = pca_M_strain_weeks_batchcorrday, width = 7, height = 6, dpi = 300, bg = "white")


#####################
pca_M_weeks_batchcorrday <- ggplot(pca_data, aes(x = PC1, y = PC2, color = weeks, shape = strain)) +
  geom_point(aes(text = hover_text), size = 3.5, alpha = 0.7) +
  #stat_ellipse(aes(group = as.factor(cage_id)), type = "t", level = 0.95, geom = "path", show.legend = FALSE) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "PCA-Male batchcorrday: weeks & strain (n=116)",
    subtitle = paste("based on", n_nona, "proteins out of", n_original),
    x = paste0("PC1 (", pca_var_perc[1], "%)"),
    y = paste0("PC2 (", pca_var_perc[2], "%)")
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(pca_M_weeks_batchcorrday)
plotly::ggplotly(pca_M_weeks_batchcorrday)

ggsave(file.path("doc", "pca_M_weeks_batchcorrday.png"), 
       plot = pca_M_weeks_batchcorrday, width = 7, height = 6, dpi = 300, bg = "white")
```

# KEGG
```{r}
library(clusterProfiler)
library(KEGGREST)
library(dplyr)

# 1) Symbols → UniProt (mouse)
gene_symbols <- rownames(df_male)
cat("Original Gene Symbols:\n"); print(head(gene_symbols))

id_conversion <- bitr(
  gene_symbols,
  fromType = "SYMBOL",
  toType   = "UNIPROT",
  OrgDb    = "org.Mm.eg.db"
)

cat("Converted IDs (Symbol → UniProt):\n"); print(head(id_conversion))

# Deduplicate / drop NAs for safety
uniprot_ids <- unique(na.omit(id_conversion$UNIPROT))

# 2) UniProt → KEGG gene (mmu:*). KEGG expects KEGG gene IDs for pathway linking.
# Build "uniprot:XXXX" query terms as KEGG likes them
kegg_up <- paste0("uniprot:", uniprot_ids)

# KEGG REST can be rate-limited; chunk the conversion to be gentle.
chunk_vec <- function(x, n = 200) split(x, ceiling(seq_along(x) / n))

kegg_conv_list <- lapply(chunk_vec(kegg_up, n = 200), function(chunk) {
  out <- tryCatch(keggConv("mmu", chunk), error = function(e) NULL)
  Sys.sleep(0.2)  # be polite to the API
  out
})

kegg_conv <- do.call(c, kegg_conv_list)

if (is.null(kegg_conv) || length(kegg_conv) == 0) {
  stop("No KEGG gene IDs were found for the provided UniProt IDs (organism 'mmu').")
}

# keggConv returns a named vector: names = "uniprot:XXXX", values = "mmu:XXXXX"
kegg_gene_ids <- unname(kegg_conv)

# 3) KEGG gene → Pathways
pathway_links <- tryCatch(
  keggLink("pathway", kegg_gene_ids),
  error = function(e) {
    message("An error occurred, possibly because no pathways were found for the given IDs.")
    return(NULL)
  }
)

if (is.null(pathway_links) || length(pathway_links) == 0) {
  stop("No KEGG pathways were found for the provided protein IDs.")
}

# 4) Build results table (map back to UniProt)
# pathway_links is a named vector: names = "mmu:XXXXX", values = "path:mmuXXXXX"
mmu_to_uniprot <- setNames(names(kegg_conv), kegg_conv)  # map "mmu:XXXXX" -> "uniprot:XXXX"

results_df <- data.frame(
  KEGG_Gene_ID     = names(pathway_links),
  KEGG_Pathway_ID  = as.character(pathway_links),
  stringsAsFactors = FALSE
) %>%
  mutate(
    UniProtID = sub("^uniprot:", "", mmu_to_uniprot[KEGG_Gene_ID]),
    KEGG_Pathway_ID = sub("^path:", "", KEGG_Pathway_ID)
  ) %>%
  select(UniProtID, KEGG_Gene_ID, KEGG_Pathway_ID) %>%
  arrange(UniProtID, KEGG_Pathway_ID)

cat("Pathways found for each UniProt ID:\n")
print(head(results_df))


```




