---
title: "Female_week12"
format: html
---

# Load dependencies
```{r setup}
#| message: false
#| warning: false

source(here::here("R/library.R"))
load(here::here("data/metadata_F.rda"))
load(here::here("data/female_log2_filt_median.rda"))
```

# functions
```{r}
# For PCA
plot_pca <- function(data, color_var, shape_var = NULL, color_vals, file_name, plot_title) {
  # Build aesthetic mapping conditionally
  mapping <- aes(
    x = PC1,
    y = PC2,
    color = .data[[color_var]],
    text = hover_text
  )
  if (!is.null(shape_var)) {
    mapping <- aes(
      x = PC1,
      y = PC2,
      color = .data[[color_var]],
      shape = .data[[shape_var]],
      text = hover_text
    )
  }
  
  p <- ggplot(data, mapping) +
    geom_point(size = 3.5, alpha = 0.7) +
    stat_ellipse(
      aes(
        group = .data[[color_var]],
        fill = .data[[color_var]]
      ),
      geom = "path",
      show.legend = FALSE
    ) +
    scale_color_manual(values = color_vals) +
    labs(
      title = plot_title,
      subtitle = paste("based on", n_nona, "proteins out of", n_original),
      x = paste0("PC1 (", pca_var_perc[1], "%)"),
      y = paste0("PC2 (", pca_var_perc[2], "%)")
    ) +
    theme_bw(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
    )
  
  print(p)
}

################################ limma main results 
# For limma
get_limma_results <- function(fit_obj, coef_name, contrast_label, alpha = 0.05) {
  
  # Extract results table
  top_table <- limma::topTable(
    fit_obj,
    coef          = coef_name,
    adjust.method = "BH",
    sort.by       = "logFC",
    number        = Inf
  ) %>%
    rownames_to_column("Protein")
  
  # Filter significant hits
  sig <- top_table %>%
    filter(adj.P.Val < alpha) %>%
    drop_na()
  
  up   <- sig %>% filter(logFC > 0)
  down <- sig %>% filter(logFC < 0)
  
  # Console summary
  message("\n---- ", contrast_label, " ----")
  message("Total significant: ", nrow(sig))
  message("UP: ", nrow(up), " | DOWN: ", nrow(down))
  
  # Return result lists
  return(list(
    top_table = top_table,
    sig       = sig,
    up        = up,
    down      = down
  ))
}


```


# Subrgoup female week12
```{r}
metadata_F <- metadata_F %>% dplyr::mutate(Group = paste(diet, strain, sep = "_"))
metadata_F12 <- metadata_F %>% dplyr::filter(weeks == "12")

df_F12 <- female_log2_filt_median[, metadata_F12$new_sample_id] # 7641
dim(df_F12) # 35 samples
all(colnames(df_F12) == metadata_F12$new_sample_id)  # Should be TRUE

metadata_F12_J <- metadata_F12 %>% dplyr::filter(strain == "J")
df_F12_J <- df_F12[, metadata_F12_J$new_sample_id]
dim(df_F12_J) # 17 samples
all(colnames(df_F12_J) == metadata_F12_J$new_sample_id)  # Should be TRUE

metadata_F12_Ntac <- metadata_F12 %>% dplyr::filter(strain == "Ntac")
df_F12_Ntac <- df_F12[, metadata_F12_Ntac$new_sample_id]
dim(df_F12_Ntac) # 18 samples
all(colnames(df_F12_Ntac) == metadata_F12_Ntac$new_sample_id)  # Should be TRUE
```
# Correlation
```{r}
all(colnames(df_F12) == metadata_F12$new_sample_id)

mat <- as.matrix(df_F12)
cor_matrix <- cor(mat, method = "pearson", use = "pairwise.complete.obs")

# 2) Annotations
annotation_col <- data.frame(
  Diet   = metadata_F12$diet,
  Strain = metadata_F12$strain,
  row.names = metadata_F12$new_sample_id
)

annotation_colors <- list(
  Diet   = c(LFD = "#21908CFF", FFMD = "#D55E00"),
  Strain = c(J = "#CCBB44", Ntac = "#AA3377")
)

F12_heatmap_corr <- pheatmap(
  cor_matrix,
  color = colorRampPalette(c("lightpink", "white", "lightskyblue3"))(100),
  main = "Pearson Correlation Heatmap Female 12week",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = TRUE,
  show_colnames = TRUE,
  scale = "none",
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  breaks = seq(0.7, 1, length.out = 100)
)

cor_values  <- cor_matrix[upper.tri(cor_matrix)]
min_corr    <- min(cor_values)
max_corr    <- max(cor_values)
median_corr <- median(cor_values)
mean_corr   <- mean(cor_values)

cat("Minimum correlation:", round(min_corr, 3), "\n")
cat("Maximum correlation:", round(max_corr, 3), "\n")
cat("Median correlation:",  round(median_corr, 3), "\n")
cat("Mean correlation:",    round(mean_corr, 3), "\n")

ggsave(file.path("doc", "F12_heatmap_corr.png"),
       plot = F12_heatmap_corr, width = 10, height = 7, dpi = 300, bg = "white")

```
# PCA data: M_12
```{r}
n_original <- nrow(df_F12) # 7642
df_nona <- na.omit(df_F12) 
n_nona <- nrow(df_nona) # 6586

# transpose data 
pca_nona <- prcomp(t(df_nona), scale = TRUE)
# quick scatter plot
factoextra::fviz_pca_ind(pca_nona) 
plot(pca_nona$x[,1], pca_nona$x[,2])

# Eigenvalues (variance explained by each PC)
pca_var <- pca_nona$sdev^2 
pca_var_perc <- round(pca_var/sum(pca_var)*100, digits = 1)
fviz_eig(pca_nona, addlabels = TRUE)  # scree plot 

pca_results <- as.data.frame(pca_nona$x) %>%
  tibble::rownames_to_column("new_sample_id")  # keep sample IDs

pca_data <- pca_results %>%
  dplyr::select(new_sample_id, PC1, PC2, PC3, PC4) %>% 
  dplyr::inner_join(metadata_phos_F_filtered, by = "new_sample_id") %>%
  dplyr::mutate(
    hover_text = paste0(
      "Sample: ", sample_id,
      "<br>Diet: ", diet,
      "<br>Strain: ", strain
    )
  )
# Should be all TRUE / empty set
all(pca_results$new_sample_id %in% metadata_phos_F12$new_sample_id)



ggsave(file.path("doc", "pca_diet_F12.png"), plot = pca_diet_F12, width = 6, height = 5, dpi = 300, bg = "white")
ggsave(file.path("doc", "pca_strain_F12.png"), plot = pca_strain_F12, width = 6, height = 5, dpi = 300, bg = "white")
```


# Limma Main Effect (additive model)
More power because all mice contribute to each main‐effect estimate.
```{r}
metadata_F12 <- metadata_F12 %>%
  dplyr::mutate(
    diet   = factor(diet, levels = c("LFD", "FFMD")),
    strain = factor(strain, levels = c("J", "Ntac"))
  )
levels(metadata_F12$diet)
levels(metadata_F12$strain)

table(metadata_F12$diet)   # FFMD-24, LFD-12
table(metadata_F12$strain) # J-17, Ntac-18

# create design matrix
design_matrix <- model.matrix(~ diet + strain + beatbox_batch, data = metadata_F12)
colnames(design_matrix)
fit  <- limma::lmFit(df_F12, design_matrix)
fit2 <- limma::eBayes(fit)

# ---- Usage ----
limma_results_F12_diet   <- get_limma_results(fit2, "dietFFMD",   "F12: Diet")
limma_results_F12_strain <- get_limma_results(fit2, "strainNtac", "F12: Strain")

# Diet histogram
hist(limma_results_F12_diet$top_table$P.Value,   main = "F12: Diet main effect",   xlab = "P-value", col = "darkslategray4", breaks = 30)
hist(limma_results_F12_strain$top_table$P.Value, main = "F12: Strain main effect", xlab = "P-value", col = "darkslategray4", breaks = 30)

png(file.path("doc", "F12_main_diet_hist.png"), width = 5, height = 4, units = "in", res = 300, bg = "white")
hist(limma_results_F12_diet$top_table$P.Value, main = "F12: Diet main effect", xlab = "P-value", col = "darkslategray4", breaks = 30)
dev.off()

png(file.path("doc", "F12_main_strain_hist.png"), width = 5, height = 4, units = "in", res = 300, bg = "white")
hist(limma_results_F12_strain$top_table$P.Value, main = "F12: Strain main effect", xlab = "P-value", col = "darkslategray4", breaks = 30)
dev.off()

```

# Limma Interaction
*dietFFMD:strainNtac* tests whether the effect of FFMD vs LFD depends on strain.
```{r}
design_interaction <- model.matrix(~ diet * strain + beatbox_batch, data = metadata_F12)
colnames(design_interaction)

fit_inter  <- limma::lmFit(df_F12, design_interaction)
fit2_inter <- limma::eBayes(fit_inter)
colnames(fit2_inter$coefficients)

limma_results_F12_interaction <- get_limma_results(
  fit2_inter,
  coef_name      = "dietFFMD:strainNtac",
  contrast_label = "F12: Diet x Strain Interaction",
  alpha          = 0.1
)

# p-value histogram
png(file.path("doc", "F12_interaction_hist.png"),
    width = 5, height = 4, units = "in", res = 300, bg = "white")
hist(
  limma_results_F12_interaction$top_table$P.Value,
  main = "F12: Interaction (diet x strain)",
  xlab = "P-value",
  col  = "darkslategray4"
)
dev.off()

# 0 hits
```

 
# Limma Contrasts
concistent with prelimiray results - the diet effect is more exaggerated in J mice than in Ntac in the short-term.
```{r}
table(metadata_F12$Group)
metadata_F12$Group <- factor(metadata_F12$Group, levels = c("LFD_J", "LFD_Ntac", "FFMD_J", "FFMD_Ntac"))
levels(metadata_F12$Group)

design_group <- model.matrix(~ 0 + Group + beatbox_batch, data = metadata_F12)
colnames(design_group)
colnames(design_group)[1:4] <- colnames(design_group)[1:4] |>
  sub("^Group", "", x = _)

contrast_matrix <- makeContrasts(
  Diet_in_J       = FFMD_J     - LFD_J,
  Diet_in_Ntac    = FFMD_Ntac  - LFD_Ntac,
  Strain_in_LFD   = LFD_Ntac   - LFD_J,
  Strain_in_FFMD  = FFMD_Ntac  - FFMD_J,
  Interaction     = (FFMD_Ntac - LFD_Ntac) - (FFMD_J - LFD_J),
  levels = design_group
)

fit  <- limma::lmFit(df_F12, design_group)
fit2 <- limma::contrasts.fit(fit, contrast_matrix)
fit2 <- limma::eBayes(fit2)

get_limma_results_contrast <- function(fit_obj, coef_name, label, alpha = 0.05) {
  top_table <- limma::topTable(
    fit_obj,
    coef          = coef_name,
    adjust.method = "BH",
    sort.by       = "logFC",
    number        = Inf
  ) %>%
    tibble::rownames_to_column("Protein")
  
  sig  <- top_table %>% dplyr::filter(adj.P.Val < alpha) %>% tidyr::drop_na()
  up   <- sig %>% dplyr::filter(logFC > 0)
  down <- sig %>% dplyr::filter(logFC < 0)
  
  # Print summary
  cat("\n----", label, "----\n")
  cat("Total significant:", nrow(sig), "\n")
  cat("Up:", nrow(up), " | Down:", nrow(down), "\n")
  
  # Save histogram if you want
  png(
    file.path("doc", paste0(label, "_hist.png")),
    width  = 5,
    height = 4,
    units  = "in",
    res    = 300,
    bg     = "white"
  )
  hist(
    top_table$P.Value,
    main = paste(label, "P-value histogram"),
    xlab = "P-value",
    col  = "darkslategray4"
  )
  dev.off()
  
  # Return results as list
  return(list(
    top_table = top_table,
    sig       = sig,
    up        = up,
    down      = down
  ))
}

# 2. Apply to All Contrasts in a Loop
# Get the contrast names you defined:
contrast_names <- colnames(contrast_matrix)

# Loop through all contrasts:
results_list <- list()
for (cname in contrast_names) {
  results_list[[cname]] <- get_limma_results_contrast(
    fit2,
    coef_name = cname,
    label     = cname,
    alpha     = 0.05  # or your chosen FDR threshold
  )
}

# results_list will store each contrast’s full results for later inspection.
# Plots and printouts are created for each contrast automatically.
limma_results_F12_contrasts <- results_list

diet_effect_J_table <- limma_results_F12_contrasts$Diet_in_J$top_table

```
# Diet effect in strains
## pca-J
```{r}
## pca-J (Female)
dim(df_F12_J) # 18 samples
df_nona <- na.omit(df_F12_J) 

n_original <- nrow(df_F12_J) # 7642
n_nona <- nrow(df_nona) # 6586

# transpose data 
pca_nona <- prcomp(t(df_nona), scale = TRUE)
# quick scatter plot
factoextra::fviz_pca_ind(pca_nona) 
plot(pca_nona$x[,1], pca_nona$x[,2])

# Eigenvalues (variance explained by each PC)
pca_var <- pca_nona$sdev^2 
pca_var_perc <- round(pca_var/sum(pca_var)*100, digits = 1)
fviz_eig(pca_nona, addlabels = TRUE)  # scree plot 

pca_results <- as.data.frame(pca_nona$x) %>%
  tibble::rownames_to_column("new_sample_id")  # keep sample IDs

pca_data <- pca_results %>%
  dplyr::select(new_sample_id, PC1, PC2, PC3, PC4) %>% 
  dplyr::inner_join(metadata_F12_J, by = "new_sample_id") %>%
  dplyr::mutate(
    hover_text = paste0(
      "Sample: ", sample_id,
      "<br>Diet: ", diet,
      "<br>Strain: ", strain
    )
  )

# Should be all TRUE / empty set
all(pca_results$new_sample_id %in% metadata_F12$new_sample_id)

pca_diet_F12_J <- plot_pca(
  data       = pca_data,
  color_var  = "diet",
  shape_var  = NULL,
  color_vals = c("LFD" = "#21908CFF", "FFMD" = "#D55E00"),
  file_name  = "pca_F12_diet_J.png",
  plot_title = "Female, 12 weeks, J (n=18)"
)

ggsave(file.path("doc", "pca_F12_diet_J.png"),
       plot = pca_diet_F12_J, width = 6, height = 5, dpi = 300, bg = "white")
```

## pca-Ntac
```{r}
## pca-Ntac (Female)
dim(df_F12_Ntac) # 18 samples
df_nona <- na.omit(df_F12_Ntac) 

n_original <- nrow(df_F12_Ntac) # 7642
n_nona <- nrow(df_nona) # 6586

# transpose data 
pca_nona <- prcomp(t(df_nona), scale = TRUE)
# quick scatter plot
factoextra::fviz_pca_ind(pca_nona) 
plot(pca_nona$x[,1], pca_nona$x[,2])

# Eigenvalues (variance explained by each PC)
pca_var <- pca_nona$sdev^2 
pca_var_perc <- round(pca_var/sum(pca_var)*100, digits = 1)
factoextra::fviz_eig(pca_nona, addlabels = TRUE)  # scree plot 

pca_results <- as.data.frame(pca_nona$x) %>%
  tibble::rownames_to_column("new_sample_id")  # keep sample IDs

pca_data <- pca_results %>%
  dplyr::select(new_sample_id, PC1, PC2, PC3, PC4) %>% 
  dplyr::inner_join(metadata_F12_Ntac, by = "new_sample_id") %>%
  dplyr::mutate(
    hover_text = paste0(
      "Sample: ", sample_id,
      "<br>Diet: ", diet,
      "<br>Strain: ", strain
    )
  )

# Should be all TRUE / empty set
all(pca_results$new_sample_id %in% metadata_F12_Ntac$new_sample_id)

pca_diet_F12_Ntac <- plot_pca(
  data       = pca_data,
  color_var  = "diet",
  shape_var  = NULL,
  color_vals = c("LFD" = "#21908CFF", "FFMD" = "#D55E00"),
  file_name  = "pca_F12_diet_Ntac.png",
  plot_title = "Female, 12 weeks, Ntac (n=18)"
)

ggsave(file.path("doc", "pca_F12_diet_Ntac.png"),
       plot = pca_diet_F12_Ntac, width = 6, height = 5, dpi = 300, bg = "white")


```


## Euler for diet
```{r}
# euler for total
diet_J_sig    <- limma_results_F12_contrasts$Diet_in_J$sig$Protein   # e.g. counts will differ
diet_Ntac_sig <- limma_results_F12_contrasts$Diet_in_Ntac$sig$Protein

# Euler diagram
fit_euler <- euler(list(J = diet_J_sig, Ntac = diet_Ntac_sig))

euler_total <- plot(
  fit_euler, quantities = TRUE, labels = TRUE,
  fills = c("#CCBB44", "#AA3377"), edges = TRUE,
  main  = "F12 Sig diet hits: total"
)

print(euler_total)
ggsave(file.path("doc", "euler_F12_Diet_effect_total.png"),
       plot = euler_total, width = 4, height = 3, dpi = 300, bg = "white") 

cat("Number unique to Ntac:", length(setdiff(diet_Ntac_sig, diet_J_sig)), "\n")
cat("Number unique to J:",    length(setdiff(diet_J_sig, diet_Ntac_sig)), "\n") 
cat("Number overlap (both):", length(intersect(diet_Ntac_sig, diet_J_sig)), "\n")
cat("Total significant (Ntac):", length(diet_Ntac_sig), "\n")
cat("Total significant (J):",    length(diet_J_sig), "\n")

#######################################################################
# euler for upregulated proteins in J & Ntac
diet_J_up    <- limma_results_F12_contrasts$Diet_in_J$up$Protein
diet_Ntac_up <- limma_results_F12_contrasts$Diet_in_Ntac$up$Protein

fit_euler_up <- euler(list(J_up = diet_J_up, Ntac_up = diet_Ntac_up))

eulerplot_up <- plot(
  fit_euler_up, quantities = TRUE, labels = TRUE,
  fills = c("#fde0dd", "#fa9fb5"), edges = TRUE,
  main  = "Significant diet hits: UP (F12)"
)

print(eulerplot_up)
ggsave(file.path("doc", "euler_F12_Diet_effect_up.png"),
       plot = eulerplot_up, width = 4, height = 3, dpi = 300, bg = "white") 

cat("Number unique to Ntac-up:", length(setdiff(diet_Ntac_up, diet_J_up)), "\n")
cat("Number unique to J-up:",    length(setdiff(diet_J_up, diet_Ntac_up)), "\n") 
cat("Number overlap (both)-up:", length(intersect(diet_Ntac_up, diet_J_up)), "\n")
cat("Total significant (Ntac)-up:", length(diet_Ntac_up), "\n")
cat("Total significant (J)-up:",    length(diet_J_up), "\n")

#####################################################################
# euler for downregulated proteins in J & Ntac
diet_J_down    <- limma_results_F12_contrasts$Diet_in_J$down$Protein
diet_Ntac_down <- limma_results_F12_contrasts$Diet_in_Ntac$down$Protein

fit_euler_down <- euler(list(J_down = diet_J_down, Ntac_down = diet_Ntac_down))

eulerplot_down <- plot(
  fit_euler_down, quantities = TRUE, labels = TRUE,
  fills = c("#ece7f2", "#2b8cbe"), edges = TRUE,
  main  = "Significant diet hits: DOWN (F12)"
)
print(eulerplot_down)

ggsave(file.path("doc", "euler_F12_Diet_effect_down.png"),
       plot = eulerplot_down, width = 4, height = 3, dpi = 300, bg = "white") 

cat("Number unique to Ntac-down:", length(setdiff(diet_Ntac_down, diet_J_down)), "\n")
cat("Number unique to J-down:",    length(setdiff(diet_J_down, diet_Ntac_down)), "\n") 
cat("Number overlap (both)-down:", length(intersect(diet_Ntac_down, diet_J_down)), "\n")
cat("Total significant (Ntac)-down:", length(diet_Ntac_down), "\n")
cat("Total significant (J)-down:",    length(diet_J_down), "\n")
```


```{r}
background_proteins <- rownames(df_F12)
length(background_proteins) # 7641

common_diet_proteins <- intersect(diet_J_sig, diet_Ntac_sig)
length(common_diet_proteins)  # 28

#mapped <- AnnotationDbi::select(org.Mm.eg.db,keys = common_diet_proteins,keytype = "SYMBOL",columns = "ENTREZID")
#entrez_ids <- unique(na.omit(mapped$ENTREZID))


ora_bp <- enrichGO(
    gene          = common_diet_proteins,
    universe      = background_proteins,
    OrgDb         = org.Mm.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 1,
    qvalueCutoff  = 0.05,
    readable      = TRUE
)


barplot(ora_bp, showCategory = 10, title = "GOBP-ORA: Common Diet-Regulated Proteins (F)")
ora_diet_F12_dotplot <- dotplot(ora_bp, showCategory = 10, title = "GOBP-ORA: Common diet Proteins (F)")
print(ora_diet_F12_dotplot)
ggsave(file.path("doc", "ora_diet_F12_dotplot.png"), plot = ora_diet_F12_dotplot, width = 6, height = 5, dpi = 300, bg = "white") 
```


## ORA -up/down
```{r}
# --- Unique UP sets (F12) ---
diet_J_up    <- limma_results_F12_contrasts$Diet_in_J$up$Protein # 52
length(diet_J_up)
diet_Ntac_up <- limma_results_F12_contrasts$Diet_in_Ntac$up$Protein # 79
length(diet_Ntac_up)

diet_J_up_uniq    <- setdiff(diet_J_up, diet_Ntac_up)  
length(diet_J_up_uniq) # 31
diet_Ntac_up_uniq <- setdiff(diet_Ntac_up, diet_J_up)  
length(diet_Ntac_up_uniq) # 58

# Background universe
background_proteins <- rownames(df_F12)  # e.g. 7612

# --- GO:BP ORA (F12) ---
ora_bp_J_up_uniq <- clusterProfiler::enrichGO(
  gene          = diet_J_up_uniq,
  universe      = background_proteins,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

ora_bp_Ntac_up_uniq <- clusterProfiler::enrichGO(
  gene          = diet_Ntac_up_uniq,
  universe      = background_proteins,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

dotplot_ora_bp_J_up_uniq    <- clusterProfiler::dotplot(ora_bp_J_up_uniq,    showCategory = 10, title = "GOBP-ORA (F12): uniq up J")
dotplot_ora_bp_Ntac_up_uniq <- clusterProfiler::dotplot(ora_bp_Ntac_up_uniq, showCategory = 10, title = "GOBP-ORA (F12): uniq up Ntac")

print(dotplot_ora_bp_J_up_uniq)
print(dotplot_ora_bp_Ntac_up_uniq)

ggsave(file.path("doc", "dotplot_ora_bp_F12_J_up_uniq.png"),
       plot = dotplot_ora_bp_J_up_uniq, width = 6, height = 6, dpi = 300, bg = "white")
ggsave(file.path("doc", "dotplot_ora_bp_F12_Ntac_up_uniq.png"),
       plot = dotplot_ora_bp_Ntac_up_uniq, width = 6, height = 6, dpi = 300, bg = "white")

# --- GO:MF ORA (F12, J uniq) ---
ora_mf_J_up_uniq <- clusterProfiler::enrichGO(
  gene          = diet_J_up_uniq,
  universe      = background_proteins,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "MF",
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

ora_mf_Ntac_up_uniq <- clusterProfiler::enrichGO(
  gene          = diet_Ntac_up_uniq,
  universe      = background_proteins,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "MF",
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

dotplot_ora_mf_J_up_uniq <- clusterProfiler::dotplot(ora_mf_J_up_uniq, showCategory = 10, title = "GOMF-ORA (F12): uniq up J")
dotplot_ora_mf_Ntac_up_uniq <- clusterProfiler::dotplot(ora_mf_Ntac_up_uniq, showCategory = 10, title = "GOMF-ORA (F12): uniq up Ntac")
print(dotplot_ora_mf_J_up_uniq)
print(dotplot_ora_mf_Ntac_up_uniq)

ggsave(file.path("doc", "dotplot_ora_mf_F12_J_up_uniq.png"),
       plot = dotplot_ora_mf_J_up_uniq, width = 6, height = 6, dpi = 300, bg = "white")
ggsave(file.path("doc", "dotplot_ora_mf_F12_Ntac_up_uniq.png"),
       plot = dotplot_ora_mf_Ntac_up_uniq, width = 6, height = 6, dpi = 300, bg = "white")

# --- GO:CC ORA (F12) ---
ora_cc_J_up_uniq <- clusterProfiler::enrichGO(
  gene          = diet_J_up_uniq,
  universe      = background_proteins,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "CC",
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

ora_cc_Ntac_up_uniq <- clusterProfiler::enrichGO(
  gene          = diet_Ntac_up_uniq,
  universe      = background_proteins,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "CC",
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

dotplot_ora_cc_J_up_uniq    <- clusterProfiler::dotplot(ora_cc_J_up_uniq,    showCategory = 10, title = "GOCC-ORA (F12): uniq up J")
dotplot_ora_cc_Ntac_up_uniq <- clusterProfiler::dotplot(ora_cc_Ntac_up_uniq, showCategory = 10, title = "GOCC-ORA (F12): uniq up Ntac")

print(dotplot_ora_cc_J_up_uniq)
print(dotplot_ora_cc_Ntac_up_uniq)

ggsave(file.path("doc", "dotplot_ora_cc_F12_J_up_uniq.png"),
       plot = dotplot_ora_cc_J_up_uniq, width = 6, height = 6, dpi = 300, bg = "white")
ggsave(file.path("doc", "dotplot_ora_cc_F12_Ntac_up_uniq.png"),
       plot = dotplot_ora_cc_Ntac_up_uniq, width = 6, height = 6, dpi = 300, bg = "white")

```


## Scatter plot Ntac vs J
```{r}
fc_J_vec <- limma_results_F12_contrasts$Diet_in_J$top_table$logFC
names(fc_J_vec) <- limma_results_F12_contrasts$Diet_in_J$top_table$Protein
fc_Ntac_vec <- limma_results_F12_contrasts$Diet_in_Ntac$top_table$logFC
names(fc_Ntac_vec) <- limma_results_F12_contrasts$Diet_in_Ntac$top_table$Protein
common_proteins <- intersect(names(fc_J_vec), names(fc_Ntac_vec))

# --- Extract significant sets ---
sig_J    <- limma_results_F12_contrasts$Diet_in_J$sig$Protein
sig_Ntac <- limma_results_F12_contrasts$Diet_in_Ntac$sig$Protein

# --- Combine into a single dataframe for plotting ---
plot_df <- data.frame(
  Protein     = common_proteins,
  logFC_J     = fc_J_vec[common_proteins],
  logFC_Ntac  = fc_Ntac_vec[common_proteins],
  stringsAsFactors = FALSE
)

# --- Category assignment ---
plot_df$sig_cat <- "Not sig"
plot_df$sig_cat[plot_df$Protein %in% sig_J    & !(plot_df$Protein %in% sig_Ntac)] <- "Sig in J only"
plot_df$sig_cat[plot_df$Protein %in% sig_Ntac & !(plot_df$Protein %in% sig_J)]    <- "Sig in Ntac only"
plot_df$sig_cat[plot_df$Protein %in% sig_J    &  plot_df$Protein %in% sig_Ntac]   <- "Sig in both"

# (Optional) enforce a consistent legend/order
plot_df$sig_cat <- factor(
  plot_df$sig_cat,
  levels = c("Sig in both", "Sig in J only", "Sig in Ntac only", "Not sig")
)

# --- Color scheme for plotting ---
plot_colors <- c(
  "Sig in both"      = "red",
  "Sig in J only"    = "orange",
  "Sig in Ntac only" = "royalblue2",
  "Not sig"          = "grey70"
)

proteins_to_label <- limma_results_F12_contrasts$Interaction$sig$Protein

cor_coeff <- cor(plot_df$logFC_J, plot_df$logFC_Ntac, method = "pearson", use = "pairwise.complete.obs")
cor_label <- paste("Pearson r =", round(cor_coeff, 2))

scatter_plot_diet <- ggplot(plot_df, aes(x = logFC_J, y = logFC_Ntac, color = sig_cat)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  scale_color_manual(values = plot_colors) +
  labs(
    x = "logFC (Diet effect, J)",
    y = "logFC (Diet effect, Ntac)",
    color = "Significance Category",
    title = "Scatter Plot (Female, n=112): Diet logFC in J vs Ntac"
  ) +
  annotate(
    "text",
    x = min(plot_df$logFC_J, na.rm = TRUE),
    y = max(plot_df$logFC_Ntac, na.rm = TRUE),
    label = cor_label,
    hjust = 0, vjust = 1, size = 5, fontface = "bold", color = "black"
  ) +
  ggrepel::geom_text_repel(
    data = subset(plot_df, Protein %in% proteins_to_label),
    aes(label = Protein),
    size = 4, box.padding = 5,
    segment.color = "black", stroke = 2, max.overlaps = Inf
  ) +
  theme_classic()

print(scatter_plot_diet)

ggsave(file.path("doc", "scatter_plot_diet2_F112.png"),
       plot = scatter_plot_diet, width = 7, height = 5, dpi = 300, bg = "white")


```

## Heatmap for common diet proteins
```{r}
diet_J_sig  <- limma_results_F12_contrasts$Diet_in_J$sig$Protein # 73
diet_Ntac_sig <- limma_results_F12_contrasts$Diet_in_Ntac$sig$Protein # 152

common_diet_proteins <- intersect(diet_J_sig, diet_Ntac_sig) # 16
print(common_diet_proteins)

# selecet proteins to show in heatmap
selected_sites <- common_diet_proteins

# find selected sites in data & filter
sites_in_data <- selected_sites[selected_sites %in% rownames(df_F12)]
filt_selected <- df_F12[sites_in_data, , drop=FALSE]

# --- 4. Create the annotation for columns (samples) ---
ann_col <- data.frame(
  Strain = metadata_F12$strain,  # Or whatever column name in your metadata for strain
  Diet   = metadata_F12$diet     # And for diet. You can include either, or both for richer annotation
)
rownames(ann_col) <- metadata_F12$new_sample_id

heatmap_common_proteins_diet <- pheatmap(
  as.matrix(filt_selected),
  annotation_col          = ann_col,    # Adds the annotation bar for strain and diet
  show_rownames           = TRUE,
  show_colnames           = TRUE,
  clustering_distance_rows= "euclidean",
  clustering_distance_cols= "euclidean",
  clustering_method       = "complete",
  scale                   = "row",      # Z-score across proteins (rows) for relative comparison
  main                    = "Heatmap of Common common diet response Proteins",
  na_col                  = "white"     # NAs will display as white
)

ggsave(file.path("doc", "heatmap_common_proteins_diet.png"),plot = heatmap_common_proteins_diet, width = 7, height = 5, dpi = 300, bg = "white")
```


# Strain difference in baseline & FFMD
```{r}
metadata_F12_lfd <- metadata_F12 %>% dplyr::filter(diet == "LFD")
df_F12_lfd <- df_F12[, metadata_F12_lfd$new_sample_id]
dim(df_F12_lfd) # 12 samples
all(colnames(df_F12_lfd) == metadata_F12_lfd$new_sample_id)  # Should be TRUE

metadata_F12_ffmd <- metadata_F12 %>% dplyr::filter(diet == "FFMD")
df_F12_ffmd <- df_F12[, metadata_F12_ffmd$new_sample_id]
dim(df_F12_ffmd) # 23 samples
all(colnames(df_F12_ffmd) == metadata_F12_ffmd$new_sample_id)  # Should be TRUE

```


## pca-LFD
```{r}
dim(df_F12_lfd) # 12 samples
df_nona <- na.omit(df_F12_lfd) 

n_original <- nrow(df_F12_lfd) # 7642
n_nona <- nrow(df_nona) # 6586

# transpose data 
pca_nona <- prcomp(t(df_nona), scale = TRUE)
# quick scatter plot
factoextra::fviz_pca_ind(pca_nona) 
plot(pca_nona$x[,1], pca_nona$x[,2])

# Eigenvalues (variance explained by each PC)
pca_var <- pca_nona$sdev^2 
pca_var_perc <- round(pca_var/sum(pca_var)*100, digits = 1)
fviz_eig(pca_nona, addlabels = TRUE)  # scree plot 

pca_results <- as.data.frame(pca_nona$x) %>%
  tibble::rownames_to_column("new_sample_id")  # keep sample IDs

pca_data <- pca_results %>%
  dplyr::select(new_sample_id, PC1, PC2, PC3, PC4) %>% 
  inner_join(metadata_F12_lfd, by = "new_sample_id") %>%
  mutate(
    hover_text = paste0(
      "Sample: ", sample_id,
      "<br>Diet: ", diet,
      "<br>Strain: ", strain
    )
  )

# Should be all TRUE / empty set
all(pca_results$new_sample_id %in% metadata_F12_lfd$new_sample_id)

pca_diet_F12_lfd <- plot_pca(
  data       = pca_data,
  color_var  = "strain",
  shape_var  = NULL,
  color_vals = c("Ntac" = "#AA3377", "J" = "#CCBB44"),
  file_name  = "pca_F12_strain_lfd.png",
  plot_title = "Female, 12 weeks, lfd (n=12)"
)

plotly::ggplotly(pca_diet_F12_lfd)

ggsave(file.path("doc", "pca_diet_F12_lfd.png"),
       plot = pca_diet_F12_lfd, width = 6, height = 5, dpi = 300, bg = "white")
```


## pca-FFMD
```{r}
dim(df_F12_ffmd) # 12 samples
df_nona <- na.omit(df_F12_ffmd) 

n_original <- nrow(df_F12_ffmd) # 7642
n_nona <- nrow(df_nona) # 6586

# transpose data 
pca_nona <- prcomp(t(df_nona), scale = TRUE)
# quick scatter plot
factoextra::fviz_pca_ind(pca_nona) 
plot(pca_nona$x[,1], pca_nona$x[,2])

# Eigenvalues (variance explained by each PC)
pca_var <- pca_nona$sdev^2 
pca_var_perc <- round(pca_var/sum(pca_var)*100, digits = 1)
fviz_eig(pca_nona, addlabels = TRUE)  # scree plot 

pca_results <- as.data.frame(pca_nona$x) %>%
  tibble::rownames_to_column("new_sample_id")  # keep sample IDs

pca_data <- pca_results %>%
  dplyr::select(new_sample_id, PC1, PC2, PC3, PC4) %>% 
  dplyr::inner_join(metadata_F12_ffmd, by = "new_sample_id") %>%
  dplyr::mutate(
    hover_text = paste0(
      "Sample: ", sample_id,
      "<br>Diet: ", diet,
      "<br>Strain: ", strain
    )
  )

# Should be all TRUE / empty set
all(pca_results$new_sample_id %in% metadata_F12_ffmd$new_sample_id)

pca_diet_F12_ffmd <- plot_pca(
  data       = pca_data,
  color_var  = "strain",
  shape_var  = NULL,
  color_vals = c("Ntac" = "#AA3377", "J" = "#CCBB44"),
  file_name  = "pca_F12_strain_ffmd.png",
  plot_title = "Female, 12 weeks, ffmd (n=12)"
)

print(pca_diet_F12_ffmd)

ggsave(
  file.path("doc", "pca_diet_F12_ffmd.png"),
  plot   = pca_diet_F12_ffmd,
  width  = 6,
  height = 5,
  dpi    = 300,
  bg     = "white"
)
```

## Euler for strain
```{r}
strain_lfd_sig  <- limma_results_F12_contrasts$Strain_in_LFD$sig$Protein  # 35
strain_ffmd_sig <- limma_results_F12_contrasts$Strain_in_FFMD$sig$Protein # 463

common_strain_proteins <- intersect(strain_lfd_sig, strain_ffmd_sig) # 16
print(common_strain_proteins)
print(setdiff(strain_lfd_sig, strain_ffmd_sig))

# Euler diagram
fit_euler_strain <- euler(list(lfd = strain_lfd_sig, ffmd = strain_ffmd_sig))

p <- plot(
  fit_euler_strain,
  quantities = TRUE,
  labels     = TRUE,
  fills      = c("#21908CFF", "#D55E00"),
  edges      = TRUE,
  main       = "Significant strain hits (F12)"
)

print(p)
ggsave(
  file.path("doc", "euler_F12_Strain_effect_total.png"),
  plot = p, width = 4, height = 3, dpi = 300, bg = "white"
)

print(setdiff(strain_lfd_sig, strain_ffmd_sig))

cat("Number unique to ffmd:", length(setdiff(strain_ffmd_sig, strain_lfd_sig)), "\n")
cat("Number unique to lfd:",  length(setdiff(strain_lfd_sig,  strain_ffmd_sig)), "\n")
cat("Number overlap (both):", length(intersect(strain_ffmd_sig, strain_lfd_sig)), "\n")
cat("Total significant (ffmd):", length(strain_ffmd_sig), "\n")
cat("Total significant (lfd):",  length(strain_lfd_sig),  "\n")

######################################################################
# up
strain_lfd_up  <- limma_results_F12_contrasts$Strain_in_LFD$up$Protein  # 11
strain_ffmd_up <- limma_results_F12_contrasts$Strain_in_FFMD$up$Protein # 196

# Euler diagram
fit_euler_strain_up <- euler(list(lfd = strain_lfd_up, ffmd = strain_ffmd_up))

p_up <- plot(
  fit_euler_strain_up,
  quantities = TRUE,
  labels     = TRUE,
  fills      = c("#fde0dd", "#fa9fb5"),
  edges      = TRUE,
  main       = "Significant strain hits-UP (F12)"
)

print(p_up)
ggsave(
  file.path("doc", "euler_F12_Strain_effect_UP.png"),
  plot = p_up, width = 4, height = 3, dpi = 300, bg = "white"
)

cat("Number unique to ffmd -UP:", length(setdiff(strain_ffmd_up, strain_lfd_up)), "\n")
cat("Number unique to lfd-UP:",  length(setdiff(strain_lfd_up,  strain_ffmd_up)), "\n")
cat("Number overlap (both)-UP:", length(intersect(strain_ffmd_up, strain_lfd_up)), "\n")
cat("Total significant (ffmd)-UP:", length(strain_ffmd_up), "\n")
cat("Total significant (lfd)-UP:",  length(strain_lfd_up),  "\n")

###############################################################################
# down
strain_lfd_down  <- limma_results_F12_contrasts$Strain_in_LFD$down$Protein  # 11
strain_ffmd_down <- limma_results_F12_contrasts$Strain_in_FFMD$down$Protein # 196

# Euler diagram
fit_euler_strain_down <- euler(list(lfd = strain_lfd_down, ffmd = strain_ffmd_down))

p_down <- plot(
  fit_euler_strain_down,
  quantities = TRUE,
  labels     = TRUE,
  fills      = c("#ece7f2", "#2b8cbe"),
  edges      = TRUE,
  main       = "Significant strain hits-DOWN (F12)"
)

print(p_down)
ggsave(
  file.path("doc", "euler_F12_Strain_effect_DOWN.png"),
  plot = p_down, width = 4, height = 3, dpi = 300, bg = "white"
)

cat("Number unique to ffmd -DOWN:", length(setdiff(strain_ffmd_down, strain_lfd_down)), "\n")
cat("Number unique to lfd-DOWN:",  length(setdiff(strain_lfd_down,  strain_ffmd_down)), "\n")
cat("Number overlap (both)-DOWN:", length(intersect(strain_ffmd_down, strain_lfd_down)), "\n")
cat("Total significant (ffmd)-DOWN:", length(strain_ffmd_down), "\n")
cat("Total significant (lfd)-DOWN:",  length(strain_lfd_down),  "\n")

```

## Common limma table
```{r}
# get toptable data (F12)
lfd_limma_table  <- limma_results_F12_contrasts$Strain_in_LFD$top_table
ffmd_limma_table <- limma_results_F12_contrasts$Strain_in_FFMD$top_table

# Filter for common proteins
common_lfd_table  <- lfd_limma_table [lfd_limma_table$Protein  %in% common_strain_proteins, ]
common_ffmd_table <- ffmd_limma_table[ffmd_limma_table$Protein %in% common_strain_proteins, ]

# Combined summary
combined_common_table <- merge(
  common_lfd_table,
  common_ffmd_table,
  by       = "Protein",
  suffixes = c("_LFD", "_FFMD")
)

print(head(combined_common_table))

```

## Heatmap sig strain dif in LFD
```{r}
# Significant sets (F12)
sig_LFD <- limma_results_F12_contrasts$Strain_in_LFD$sig$Protein
print(sig_LFD)

# Select proteins to show in heatmap
selected_sites <- sig_LFD

# Find selected sites in data & filter
sites_in_data <- selected_sites[selected_sites %in% rownames(df_F12)]
filt_selected <- df_F12[sites_in_data, , drop = FALSE]

# --- Annotation for columns (samples) ---
ann_col <- data.frame(
  Strain = metadata_F12$strain,
  Diet   = metadata_F12$diet
)
rownames(ann_col) <- metadata_F12$new_sample_id

# Heatmap
heatmap_straineffect_sighits_LFD <- pheatmap(
  as.matrix(filt_selected),
  annotation_col           = ann_col,
  show_rownames            = TRUE,
  show_colnames            = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method        = "complete",
  scale                    = "row",
  main                     = "Heatmap of sig strain dif hits in LFD (F12)",
  na_col                   = "white"
)

ggsave(
  file.path("doc", "heatmap_straineffect_sighits_LFD.png"),
  plot   = heatmap_straineffect_sighits_LFD,
  width  = 7, height = 4, dpi = 300, bg = "white"
)

```


## Heatmap sig strain dif in FFMD
```{r}
# Significant sets (F12)
sig_FFMD <- limma_results_F12_contrasts$Strain_in_FFMD$sig$Protein
print(sig_FFMD)

# Select proteins to show in heatmap
selected_sites <- sig_FFMD

# Find selected sites in data & filter
sites_in_data <- selected_sites[selected_sites %in% rownames(df_F12)]
filt_selected <- df_F12[sites_in_data, , drop = FALSE]

# --- Annotation for columns (samples) ---
ann_col <- data.frame(
  Strain = metadata_F12$strain,
  Diet   = metadata_F12$diet
)
rownames(ann_col) <- metadata_F12$new_sample_id

# Heatmap
heatmap_straineffect_sighits_FFMD<- pheatmap(
  as.matrix(filt_selected),
  annotation_col           = ann_col,
  show_rownames            = TRUE,
  show_colnames            = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method        = "complete",
  scale                    = "row",
  main                     = "Heatmap of sig strain dif hits in FFMD (F12)",
  na_col                   = "white"
)

ggsave(
  file.path("doc", "heatmap_straineffect_sighits_FFMD.png"),
  plot   = heatmap_straineffect_sighits_FFMD,
  width  = 7, height = 5, dpi = 300, bg = "white"
)
```

